Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2014-04-11 pituba/article/article.tex
more and more and more.

\begin{document}
\begin{frontmatter}
\title{Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equations}
%\title{Obtaining the classical material balance equations from black-oil equations}
%\journal{Journal of Petroleum Science and Engineering}
%\author{R.B. Piccinini}\ead{rbpiccinini@gmail.com}
%\address{Petrobras - UO-BA/ATP-N/RES. Salvador, Bahia, Brazil.}

\begin{abstract}
This work presents a derivation of the classical material balance equations starting from black-oil equations in integral form. The existing link between both approaches is discussed and an interpretation for pressure in material balance calculations is provided. It is shown that a condition stronger than hydrostatic equilibrium is needed to reduce black-oil equations to the classical approach. The classical equations are formulated for a variable bubble-point pressure, extending their applicability to reservoirs under pressure maintenance. Sufficient conditions for solution uniqueness are presented and a scheme for numerical solution is proposed.
\end{abstract}

@@

\section{Introduction}
The so called material balance equations are widely employed in reservoir engineering and have become a classical tool in oil reservoir management. The equations were originally presented in \cite{schilthuis1936active} and conceived as a zero dimensional model for estimating reservoir pressure from a balance of fluid production and injection. \cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the produced volume would come from and where pressure would present a similar behavior. Such regions were said to contain an \textit{active oil}.

The equations developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, phase or thermal equilibrium. Also, the zero-dimensional approach excluded the possibility of mechanical equilibrium under gravity and capillary forces, which produces a variable fluid distribution in space. An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}. The novelty on hydrodynamics was the introduction of a macroscopic equation of motion employing Darcy's law.

Black-oil equations arise from the generalization of the zero dimensional concept to a multidimensional macroscopic model which incorporates hydrodynamics via Darcy's law and entails exactly the same thermodynamic approach for phase equilibrium \citep{aziz1979petroleum}.

%Although many applications of material balance equations have been reported, few explore their connection.

The term black-oil refers to specific hydrocarbon compositions whose phase equilibrium for isothermal changes in pressure may be well described by a two pseudo-component system. For such reservoir fluids, gas and oil compositions remain essentially constant in the entire range of pressure observed during reservoir productive life.

@@

Since black-oil and material balance equations differ only on spatial scale and the existence of local pressure gradients, there must exist conditions under which material balance equations might be recovered from the black-oil approach. This work examines such conditions.

This existing link was briefly addressed in \cite{ertekin2001basic}, where both approaches were argued to be equivalent if capillarity and gradients of flow potential $\Phi$ could be neglected, i.e.:
\begin{subequations}
\begin{equation}
p_\alpha = p \, ,
\end{equation}
\begin{equation}
\nabla \Phi_\alpha = \nabla p_\alpha - \rho_\alpha g \nabla z = 0 \, ,
\end{equation}
\end{subequations}
for $\alpha=o,g,w$ referring to oil, gas and water phases.

@@

The present derivation of material balance equations allows bubble-point pressure to vary with time accordingly to the same concept widely employed in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}. The employed approach extends material balance equations to saturated reservoirs under pressure maintenance (e.g. fluid injection). The applicability, however, depend on how important non-equilibrium effects are to the phase equilibrium.

@@

This section starts presenting black-oil equations and it proceeds averaging quantities in space and examining simplifying assumptions for wich material balance equations are obtained.

@@

Another simplification is now made assuming thermal equilibrium in the entire reservoir volume:
\begin{equation}
T_w = T_o = T_g = T
\end{equation}
and
\begin{equation}
\nabla T \equiv 0 \, .
\end{equation}

@@

Derivation of the classical material balance equations consists in assuming no fluid motion and in integrating black-oil equations in the entire reservoir volume. The integration process, however, requires all quantities to be either averaged or assumed constant in space.

@@

We now define space averaged quantities for porosity ($\phi$), saturation ($S_\alpha$) and density ($\rho_\alpha$) of a phase $\alpha$ and mass concentration of a component $k$ in phase $\alpha$ ($Y_{\alpha,k}$).

@@

Applications of material balance equations are widely employed to stablish a relation between production (and injection) and reservoir pressure. It turns out, as we know, that pressure is different for different locations, mostly for different values of depth, and also for different phases.

What is then the significance of a single value of pressure for the entire reservoir, i.e., the reservoir pressure?

We shall see that for the purpose of computing phase densities in the material balance, if we define a reference depth and use the corresponding local pressure, no important error is introduced in computing density. Phase pressure increases with depth according to the hydrostatic gradient. The pressure increase, however, is not significant to produce appreciable variation in phase density.

@@

We may quantify this approximation by putting together the expressions for fluid compressibility and hydrostatic equilibrium.
For this, we assume a constant isothermal compressibility for oil phase
\begin{equation}\label{eq: grado}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o 
\end{equation}
and we use hydrostatic equilibrium $dp_o = \rho_o g dz$ to obtain the expression for the derivative of oil density with depth:
\begin{equation}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
\end{equation}

@@

For the case of an oil reservoir covered by a gas cap, the reference depth may be defined with advantage at the gas-oil contact. The point is that at the contact, both gas and oil densities may be computed using the same pressure, i.e., the corresponding local value.

@@

\subsubsection{Material balance as a pressure equation}

@@

Figure~\ref{fig: epsilon} shows the shape of residual function ($\epsilon$) with respect to pressure, i.e, holding all produced or injected volumes constant and allowing only pressure to change. Each different solid line corresponds to a different volume of fluid production and injection. The function root is the solution of material balance equation, Eq.~\eqref{eq: MBE_short}, and the non-differentiable point is the bubble-point pressure. Also, it is seen that function $\epsilon$ monotonically decreases with pressure, as theoretically shown before.

>>>>

\begin{document}
\begin{frontmatter}
\title{Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equation}
%\title{Obtaining the classical material balance equations from black-oil equations}
%\journal{Journal of Petroleum Science and Engineering}
%\author{R.B. Piccinini}\ead{rbpiccinini@gmail.com}
%\address{Petrobras - UO-BA/ATP-N/RES. Salvador, Bahia, Brazil.}

\begin{abstract}
This work presents a derivation of the classical material balance equation starting from black-oil equations in integral form. The existing link between both approaches is examined in detail, including the necessary step of averaging PVT properties. An interpretation for pressure in material balance calculations is provided and the choice of a reference depth (datum) is discussed. The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented. A scheme for numerical solution is also proposed.
\end{abstract}

@@

\section{Introduction}
The so called material balance equation is widely employed in reservoir engineering and has become a classical tool in oil reservoir management. The equation was originally presented in \cite{schilthuis1936active} and conceived as a zero dimensional model for estimating reservoir pressure from a balance of fluid production and injection. \cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the produced volume would come from and where pressure would present a similar trend. Such regions were said to contain an \textit{active oil}.

The equations developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, phase or thermal equilibrium. Also, the zero-dimensional approach excluded the effects of mechanical equilibrium under gravity and capillary forces and the resulting phase segregation. An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}. The novelty on hydrodynamics was the introduction of a macroscopic equation of motion employing Darcy's law.

Black-oil equations arise from the generalization of the zero dimensional concept to a three-dimensional macroscopic model which incorporates Darcy's law while entailing the same approach for phase equilibrium \citep{aziz1979petroleum}.

The term black-oil refers to specific hydrocarbon compositions whose phase equilibrium for isothermal changes in pressure may be well described by a two pseudo-component system. For such reservoir fluids, compositions of gas and oil phases remain essentially constant in the entire range of pressure observed during production life.

@@

%Since black-oil and material balance equations differ only on spatial scale and the existence of local pressure gradients, there must exist conditions under which material balance equations might be recovered from the black-oil approach. This work examines such conditions.

Although many applications of material balance equation have been reported in the literature, few explore its connection with black-oil equations. \cite{ertekin2001basic} briefly addressed this issue arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \mathbf{g}$) could be neglected and if PVT properties were averaged.

%\begin{subequations}
%\begin{equation}
%p_\alpha = p \, ,
%\end{equation}
%\begin{equation}
%\nabla \Phi_\alpha = \nabla p_\alpha - \rho_\alpha g \nabla z = 0 \, ,
%\end{equation}
%\end{subequations}
%for $\alpha=o,g,w$ referring to oil, gas and water phases.

@@

The present derivation of material balance equations allows bubble-point pressure to vary accordingly to the same concept widely employed in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}. The employed approach extends material balance equations to saturated reservoirs under pressure maintenance (e.g. fluid injection). The applicability, however, depend on how important non-equilibrium effects are to the phase equilibrium.

@@

This section starts presenting black-oil equations and it proceeds averaging quantities in space and examining simplifying assumptions for which material balance equations are obtained.

@@

Another simplification is now made assuming thermal equilibrium in the entire reservoir volume:
\begin{equation}
T_w = T_o = T_g = T_{rock} = T
\end{equation}
and
\begin{equation}
\nabla T \equiv 0 \, .
\end{equation}

@@

Derivation of the material balance equation consists in assuming no fluid motion and in integrating black-oil equations in reservoir volume. The integration process, however, requires all quantities to be either spatially averaged or assumed constant.

@@

We now define spatially averaged quantities for porosity ($\phi$), saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration of a component $k$ in phase $\alpha$ ($Y_{\alpha,k}$).

@@

Material balance equation is widely employed to stablish a relation between production (and injection) and reservoir pressure. It turns out, as we know, that pressure is different for different locations, mostly for different values of depth, and also for different phases.

What is then the significance of a single value of pressure for the entire reservoir?

We shall see that for the purpose of computing the averaged value of phase densities in material balance, we may define some reference depth and use the corresponding local pressure without introducing any important error. Phase pressure increases with depth according to the hydrostatic gradient. The pressure increase, however, is not significant to produce appreciable variation in density.

@@

We may quantify this approximation by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase
\begin{equation}\label{eq: grado}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o 
\end{equation}
and we use hydrostatic equilibrium $dp_o = \rho_o g dz$ to obtain the expression for the derivative of oil density with depth:
\begin{equation}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
\end{equation}

@@

For the case of an oil reservoir covered by a gas cap, the reference depth may be defined, with advantage, at the gas-oil contact. The point is that at the contact both gas and oil densities may be computed using the same pressure, i.e., the corresponding local value.

@@

\subsubsection{Material balance as an equation for reservoir pressure}

@@

Figure~\ref{fig: epsilon} shows the shape of residual function ($\epsilon$) with respect to pressure, i.e., holding all produced or injected volumes constant and allowing only pressure to change. Each different solid line corresponds to a different volume of fluid production and injection. The function root is the solution of material balance equation, Eq.~\eqref{eq: MBE_short}, and the non-differentiable point is the bubble-point pressure. Also, it is seen that function $\epsilon$ monotonically decreases with pressure, as theoretically shown before.
